---
title: "TARGETv1-VAL-20250008"
author: "Caryn Hale"
date: "`r Sys.Date()`"
output: html_document
params:
  analysis_dir: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggsci)
library(ggpubr)
library(tximport)
library(ggh4x)
library(viridis)
library(rstatix)
library(NatParksPalettes)
library(ggrepel)
library(plotly)
library(beyonce)

# wd <- "/Users/halec/crater2/dmp/dms/qc-data/TARGET/2025/TARGETv1-VAL-20250008/"
# pool <- "TARGETv1-VAL-20250008"
# samples <- list.files(wd)
analysis_root <- params$analysis_dir
sample_dirs <- list.dirs(analysis_root, full.names = TRUE, recursive = FALSE)
samples <- basename(sample_dirs)

# my_colors <- NatParksPalettes::natparks.pals("Banff", 7)[c(1,3, 4, 6, 7, 8, 2)]
# more_colors <- NatParksPalettes::natparks.pals("BryceCanyon", 9)[c(1,3, 4, 6, 7, 8, 2)]
my_colors <- beyonce_palette(18)

```

## Total Reads Sequenced

This is the # of read pairs from the uncollapsed BAM file

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}

bam_stat_results_dup <- tibble()

for (i in seq_along(samples)) {
# wds <- file.path(paste(wd, samples[i], "/DEFAULT/forte/analysis/", samples[i], sep = ""))
wds <- file.path(paste(analysis_root, samples[i], sep = ""))

dup_file <- list.files(paste(wds, "/rseqc/dup/", sep = ""))
bam_stat <- dup_file[grep("bam_stat", dup_file)]

# setwd(paste(wds, "/rseqc/dup/", sep = ""))
bam_stat_path <- file.path(wds, "rseqc", "dup", bam_stat)

bam_stat_results_dup <- bind_rows(bam_stat_results_dup,
                              read_tsv(bam_stat_path, col_names = FALSE, trim_ws = TRUE, skip = 4) %>%
  mutate(X1 = str_squish(X1)) %>%
  separate(X1, into = c("stat", "value"), sep = ": ") %>%
  dplyr::filter(is.na(value)==FALSE) %>%
  mutate(sample = samples[i]))
}

ggplotly(ggplot(data = bam_stat_results_dup %>%
         dplyr::filter(stat == "Read-1") %>%
           mutate(flag = if_else(as.numeric(value) < 50000000, "warn", "pass")),
       aes(x = sample,
           y = as.numeric(value),
           fill = flag)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_fill_manual(values = my_colors, name = "50M Reads?") +
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6, sep = "")) +
  labs(y = "Total Reads", x = ""))
```

## Unique Reads (post-deduplication)

This is the # of read pairs from the deduplicated BAM file


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
bam_stat_results_dedup <- tibble()

for (i in seq_along(samples)) {
# wds <- file.path(paste(wd, samples[i], "/DEFAULT/forte/analysis/", samples[i], sep = ""))
wds <- file.path(paste(analysis_root, samples[i], sep = ""))

dedup_file <- list.files(paste(wds, "/rseqc/dedup/", sep = ""))
bam_stat <- dedup_file[grep("bam_stat", dedup_file)]

# setwd(paste(wds, "/rseqc/dedup/", sep = ""))
bam_stat_path <- file.path(wds, "rseqc", "dedup", bam_stat)

bam_stat_results_dedup <- bind_rows(bam_stat_results_dedup,
                              read_tsv(bam_stat_path, col_names = FALSE, trim_ws = TRUE, skip = 4) %>%
  mutate(X1 = str_squish(X1)) %>%
  separate(X1, into = c("stat", "value"), sep = ": ") %>%
  dplyr::filter(is.na(value)==FALSE) %>%
  mutate(sample = samples[i]))
}

ggplotly(ggplot(data = bam_stat_results_dedup %>%
         dplyr::filter(stat == "Read-1") %>%
           mutate(flag = if_else(as.numeric(value) < 10000000, "fail", 
                                 if_else(between(as.numeric(value), 10000001, 50000000)==TRUE, "warn", "pass"))),
       aes(x = sample,
           y = as.numeric(value),
           fill = flag)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single", width = .9)) +
  theme_bw() +
  theme(plot.margin = unit(c(.5,.5,.5,.5), "cm")) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_fill_manual(values = my_colors, name = "QC flag") +
  geom_hline(yintercept = 10000000, color = "red", linetype = "dashed") +
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6, sep = "")) +
  labs(y = "Unique Reads", x = ""))
```

## Total vs Unique Reads

In order to see the amount of collapse that occurred in the deduplication (this is a combination of the two previous graphs)

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
reads_both <- bind_rows(
  bam_stat_results_dup %>%
    mutate(type = "total"),
  bam_stat_results_dedup %>%
    mutate(type = "unique")
) %>%
  dplyr::filter(stat == "Read-1")

ggplot(data = reads_both,
       aes(x = sample,
           y = as.numeric(value),
           fill = type)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single", width = .9)) +
  theme_bw() +
  scale_fill_manual(name = "read type", values = my_colors[c(1, 5)]) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  #facet_wrap(~type, scales = "free_x") +
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6, sep = "")) +
  labs(y = "reads", x = "sample")
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}
picard_dedup_results <- tibble()

for (i in seq_along(samples)) {
# wds <- file.path(paste(wd, samples[i], "/DEFAULT/forte/analysis/", samples[i], sep = ""))
wds <- file.path(paste(analysis_root, samples[i], sep = ""))

picard_dedup_file <- list.files(paste(wds, "/picard/dedup/", sep = ""))
RNAmetrics.file <- picard_dedup_file[grep("rna_metrics", picard_dedup_file)]

# setwd(paste(wds, "/picard/dedup/", sep = ""))
RNAmetrics_path <- file.path(wds, "picard", "dedup", RNAmetrics.file)

picard_dedup_results <- bind_rows(picard_dedup_results,
                              read_tsv(RNAmetrics_path,
                                       skip = 6)[1,]  %>%
                                type_convert() %>%
                                mutate(sample = samples[i]))
}

```


## Mapping Statistics

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=4}
regions <- picard_dedup_results %>%
  dplyr::select(sample, PCT_RIBOSOMAL_BASES, PCT_CODING_BASES, PCT_UTR_BASES, PCT_INTRONIC_BASES, PCT_INTERGENIC_BASES) %>%
  pivot_longer(cols = 2:6,
               names_to = "region",
               values_to = "pct") %>%
         mutate(region = gsub("PCT_", "", region)) %>%
         mutate(region = gsub("_BASES", "", region))

ggplot(data = regions,
       aes(x = sample,
           y = pct,
           fill = region)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_fill_manual(values = my_colors) +
  theme(plot.margin = unit(c(1,1,1,1.5), "cm")) +
  labs(x = "",
       y = "fraction")


```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=4}
ggplotly(ggplot(data = regions %>%
         filter(region == "CODING") %>%
         mutate(flag = if_else(pct < .8, "warn", "pass")),
       aes(x = sample,
           y = pct*100,
           fill = flag)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors, name = "QC flag") +
  theme_bw() +
  geom_hline(yintercept = 80, color = "red", linetype = "dashed") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x = "sample", y = "% coding reads", title = "Coding Reads"))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=4}
ggplotly(ggplot(data = regions %>%
         filter(region == "INTERGENIC") %>%
         mutate(flag = if_else(pct > .02, "warn", "pass")),
       aes(x = sample,
           y = pct*100,
           fill = flag)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors, name = "QC flag") +
  theme_bw() +
  geom_hline(yintercept = 2, color = "red", linetype = "dashed") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x = "sample", y = "% intergenic reads", title = "Intergenic Reads"))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=4}
ggplotly(ggplot(data = regions %>%
         filter(region == "INTRONIC") %>%
         mutate(flag = if_else(pct > .05, "warn", "pass")),
       aes(x = sample,
           y = pct*100,
           fill = flag)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors, name = "QC flag") +
  theme_bw() +
  geom_hline(yintercept = 5, color = "red", linetype = "dashed") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x = "sample", y = "% intronic reads", title = "Intronic Reads"))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=4}
ggplotly(ggplot(data = regions %>%
         filter(region == "RIBOSOMAL") %>%
         mutate(flag = if_else(pct > .001, "warn", "pass")),
       aes(x = sample,
           y = pct*100,
           fill = flag)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors, name = "QC flag") +
  theme_bw() +
  geom_hline(yintercept = .1, color = "red", linetype = "dashed") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x = "sample", y = "% ribosomal reads", title = "Ribosomal Reads"))
```


## Detected genes (tpm \> 2)

This is the number of genes detected by pseudoalignment with a deduplicated tpm > 2

```{r eval=TRUE, echo = FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, include=TRUE}
tx2gene <- read_tsv("/research/huynhn3/projects/target_qc/forte/modules/local/target_dmp_qc/Homo_sampiens_tx2gene_Ensembl_v111.tsv")

# kallisto_files <- Sys.glob(file.path(paste(wd, "*/DEFAULT/forte/analysis/*/kallisto/abundance.h5", sep = "")))
# setwd(wd)
# kallisto_files2 <- gsub(wd, "", kallisto_files)
kallisto_files <- Sys.glob(file.path(analysis_root, "*", "kallisto", "abundance.h5"))

txi <- tximport(kallisto_files, type = "kallisto", tx2gene = tx2gene, ignoreTxVersion = TRUE)

# save(txi, file = "~/Desktop/txi.RData")
# load(file = "~/Desktop/txi.RData")
```

```{r eval=TRUE, echo = FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, include=TRUE}
abundance <- data.frame(txi$abundance)
names(abundance) <- samples

abundance_current <- abundance

abundance_num_genes <- abundance %>%
  as_tibble(rownames = "Gene") %>%
  pivot_longer(2:(dim(abundance)[2] + 1), names_to = "sample", values_to = "tpm") %>%
  dplyr::filter(tpm > 2)

summary <- abundance %>%
  as_tibble(rownames = "Gene") %>%
  pivot_longer(2:(dim(abundance)[2] + 1), names_to = "sample", values_to = "tpm") %>%
  dplyr::filter(tpm > 2) %>%
  group_by(sample) %>%
  summarise(genes = n())

ggplot(data = summary,
       aes(x = sample,
           y = genes,
           fill = sample)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
  scale_fill_manual(values = my_colors) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  labs(x = "", y = "# of genes (tpm > 2)")

```


## Spliced Reads

We calculate the number of spliced reads detected. This is an indication of RNA purity and quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
spliced_reads <- bam_stat_results_dedup %>%
  dplyr::filter(stat %in% c("Non-splice reads", "Splice reads")) %>%
  mutate(value = as.numeric(value)) %>%
  pivot_wider(names_from = stat,
              values_from = value) %>%
  mutate(percent_spliced = `Splice reads` / (`Splice reads` + `Non-splice reads`))

ggplot(data = spliced_reads %>%
         mutate(flag = if_else(percent_spliced < .4, "warn", "pass")),
       aes(x = sample,
           y = percent_spliced*100,
           fill = flag)) +
  geom_col() +
  scale_fill_manual(values = my_colors, name = "QC flag") +
  labs(y = "% spliced reads", x = "") +
  theme_bw() +
  geom_hline(yintercept = 40, color = "red", linetype = "dashed") +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```
